### 2018-04-19 write-up/background.tex
Start background and add matrices to semantics

\begin{guidance}
    A more extensive coverage of what's required to understand your
    work. In general you should assume the reader has a good undergraduate
    degree in computer science, but is not necessarily an expert in
    the particular area you've been working on. Hence this chapter
    may need to summarize some ``text book'' material.

    This is not something you'd normally require in an academic paper,
    and it may not be appropriate for your particular circumstances.
    Indeed, in some cases it's possible to cover all of the ``background''
    material either in the introduction or at appropriate places in
    the rest of the dissertation.

\end{guidance}

\section{Tracking Resources with Linearity}

\section{Problem in Detail}

\section{Proposed Solution}

>>>>

% \begin{guidance}
%     A more extensive coverage of what's required to understand your
%     work. In general you should assume the reader has a good undergraduate
%     degree in computer science, but is not necessarily an expert in
%     the particular area you've been working on. Hence this chapter
%     may need to summarize some ``text book'' material.
%
%     This is not something you'd normally require in an academic paper,
%     and it may not be appropriate for your particular circumstances.
%     Indeed, in some cases it's possible to cover all of the ``background''
%     material either in the introduction or at appropriate places in
%     the rest of the dissertation.
% \end{guidance}

\prechapter{%
    In this chapter, I will outline the concept of linear types and show how
    they can be used to solve the problems faced by programmers writing code
    using linear-algebra libraries. I will be emphasising the \emph{practical}
    and intuitive explanations of linear types to keep this thesis accessible
    to working programmers as well as academics not familiar with type-theory;
    giving only a terse overview of the history and theory behind linear types
    for the interested reader to pursue further.
}%

\section{Tracking Resources with Linearity}
Familiar examples of using a type-system to express program-invariants are
existential-types for abstraction and encapsulation, polymorphic types for
parametricity and composition (a.k.a generics). Less-known examples include
dependent-types (contracts or pre- and post-conditions). The advantages of
using a type-system to express program invariants are summarised by saying the
stronger the rules you follow, the better the guarantees you can get about your
program, \emph{before} you run it. At first, the rules seems restrictive, but
similar to how the rules of grammar, spelling and more generally writing help a
writer make it easier and clearer to communicate the ideas they wish to express,
so to do rules about typing make it easier to communicate the intent and
assumptions under which a program is written. An added, but often overlooked
benefit is automated-checking: a programmer can boldly refactor in certain
ways and the compiler will \emph{assist} in ensuring the relevant invariants
the type-system enforces are updated and kept consistent by pointing-out
where they are violated.

Linear types are a way to help a programmer track and manage resources. In
practical programming terms, they enforce the restriction that a value may be
used exactly once.\footnote{This definition may differ from more colloquial
uses in discussions surrounding \emph{substructural} type systems and/or Rust.
A brief explanation can be found in Section~\ref{sec:further}.} While this
restriction may seem limiting at first, precisely these constraints can be used
to express common invariants of the programs written by working programmers
every day. For example: a file or a socket, once opened \emph{must} closed; all
memory that is manually allocated must be freed.  C++'s destructors and Rust's
Drop-trait (and more generally, its borrow-checker) attempt to enforce these
constraints by basically doing the same thing: any resource that has not been
moved is deallocated at the end of the current lexical scope. Notably, these
languages also permit aliasing, alongside rules enforcing when it is acceptable
to do so.  On face value, the above one-line description of linear types
prevents aliasing or functions such as $\lambda x.\;x \times x$, such features
are still allowed (albeit in a more restricted fashion) in a \emph{usable}
linear type system designed for working with linear-algebra libraries.

\section{Problem in Detail}
Given this background, the most pertinent question at hand is: what problems do
linear-algebra library users (and writers) typically face? The answer to this
question depends on which of two buckets a programmer falls (or is forced by
domain) into. On one side, we have users of high-level linear-algebra libraries such
as Owl (for OCaml), Julia and Numpy (for Python); other the other, we have
users of more manual, lower-level libraries such as BLAS (Basic Linear Algebra
Subroutines) and Eigen for languages like C++ and FORTRAN.\footnote{I am not including
Rust in this comparison because its linear-algebra libraries are under active
development and not as well-known/used. Later on, given that it is a language
with in-built support of substructural features to track resources, Rust will
be compared and constrasted with this project to evaluate the classic
(E)DSL-versus-language-feature debate as it applies to the domain of
linear-algebra libraries.}
% https://github.com/AtheMathmo/rulinalg/blob/e0d8b76ddf8bbb5547cb9c98172e594605ab5fef/src/matrix/impl_ops.rs
% Which operand is overwritten determines scope. Would make sense to argue
% pros-cons of implicitness vs explicitness, emulating Rust's Affine with
% linear sugar.

\subsection{One Too Many Copies and a Thousand Bytes Behind}
% CODE EXAMPLES

\subsection{IHNIWTNM}
The title of this subsection\footnote{I Have No Idea What Those Names Mean.}
illustrates the problem with the C++/FORTRAN side: legibility (and ease of
reasoning) is sacrified at the alter of performance and efficiency.

% CODE EXAMPLES

\section{Proposed Solution}
A (E)DSL!

% CODE EXAMPLES

\section{Further Reading and Theory}\label{sec:further}
\begin{itemize}
    \item History - Girard and so on
    \item Congratulations, it's a co-monad
    \item Co-effects
    \item General direction of field - Haskell, Rust, RAML, Atkey/McBride \& Idris, Krishnaswami
\end{itemize}

### 2018-04-19 write-up/evaluation.tex
Start background and add matrices to semantics

>>>>

\prechapter{%
    In this chapter, I will argue the central premise of this thesis: linear
    types are a practical and usable tool to help working programmers
    write code that is both more legible and less resource-hungry than with
    existing linear-algebra frameworks. My project illustrates how to do so in
    a way that can be implemented as a usable \emph{library} for existing
    languages and frameworks that leverages the already impressive amount of
    work gone into optimising them so far.
}%

\section{Against Overly-Safe Copying}
% Numpy, Julia, Owl

\section{Against Poor Legibility}
% C++, Fortran

### 2018-04-19 write-up/implementation.tex
Start background and add matrices to semantics

@@

Here, I will describe the syntax and semantics.

>>>>

\prechapter{%
    I implemented this project in OCaml, however I strongly believe the ideas
    described in this chapter can be applied easily to other languages and also
    are modular enough to extend the OCaml implementation to output to different
    back-end languages. I will show how a small core language with a few features
    can be the target of heavy-desugaring of typical linear-algebra programs. This
    core language can then be elaborated and checked for linearity before
    performing some simple and predictable optimisations and emitting (in this
    particular implementation) OCaml code that is not obviously safe, correct
    and performant.
}%

@@
